\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-03 paper.tex
Fix

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it has more than one child. Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\%.

@@

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess,
it is not clear that generating many guesses
when sample sizes are large will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of plausible ARGs.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference, and to
incorporate uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have more than two children (polytomies
representing uncertainty over the ordering of coalescence events, Appendix~\ref{sec-cell-lineages-and-args}) or
more than two parents (representing uncertainty over the ordering
of multiple recombination events, Section~\ref{sec-ARG-simplification}). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
% Obvious, but nobody is doing it, so needs to be said.
% TODO rephrase to say this is an open question.
Fully utilising such uncertainty in downstream
applications is also vital.

@@

% This is all a bit boring, maybe, but I think the people we want to
% influence into using tskit will need to have this spelled out for them.
Rigorously defining interchange formats~\cite[e.g.][]{kelleher2019htsget}
is difficult and time-consuming,
and no matter how precise the specification, in practise
it is the \emph{implementations} that determine how well methods
interoperate. The BAM read alignment format~\citep{li2009sequence}
is an instructive example. Originally developed as part of the
1000 Genomes project~\citep{10002015global} to address the fragmented
software ecosystem that existed at the time~\citep{danecek2021twelve},
BAM has since become ubiquitous in bioinformatics pipelines.
The excellent interoperability between methods
exchanging alignment data is largely attributable to the
success of \texttt{htslib}~\citep{bonfield2021htslib}, the
software library that \emph{implements} BAM and several other
foundational bioinformatics file formats.
Today, there are thousands of software projects
using \texttt{htslib}~\citep{bonfield2021htslib},
and it this shared use of community software
infrastructure that guarantees the smooth flow of data between
applications.
The emerging ARG software ecosystem could similarly benefit
from the adoption of such shared community infrastructure
to handle the mundane and time-consuming details of data interchange.
The \texttt{tskit} library (Section~\ref{sec-efficiency})
is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
% TODO needs rephrasing
Adopting it as a community standard may ease software implementation burden
on researchers, freeing their time to address the many fascinating open
questions and challenges that exist.

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
Peter Ralph and Andrew Vaughan
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.

>>>>

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it has more than one child. Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\%.

@@

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess,
it is not clear that generating many guesses
when sample sizes are large will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of plausible ARGs.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference, and to
incorporate uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have more than two children (polytomies
representing uncertainty over the ordering of coalescence events, Appendix~\ref{sec-cell-lineages-and-args}) or
more than two parents (representing uncertainty over the ordering
of multiple recombination events, Section~\ref{sec-ARG-simplification}). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
How this uncertainty can be utilised in downstream applications
is an open question.

@@

% This is all a bit boring, maybe, but I think the people we want to
% influence into using tskit will need to have this spelled out for them.
Rigorously defining interchange formats~\cite[e.g.][]{kelleher2019htsget}
is difficult and time-consuming,
and no matter how precise the specification, in practise
it is the \emph{implementations} that determine how well methods
interoperate. The BAM read alignment format~\citep{li2009sequence}
is an instructive example. Originally developed as part of the
1000 Genomes project~\citep{10002015global} to address the fragmented
software ecosystem that existed at the time~\citep{danecek2021twelve},
BAM has since become ubiquitous in bioinformatics pipelines.
The excellent interoperability between methods
exchanging alignment data is largely attributable to the
success of \texttt{htslib}~\citep{bonfield2021htslib}, the
software library that \emph{implements} BAM and several other
foundational bioinformatics file formats.
Today, there are thousands of software projects
using \texttt{htslib}~\citep{bonfield2021htslib},
and it this shared use of community software
infrastructure that guarantees the smooth flow of data between
applications.
The emerging ARG software ecosystem could similarly benefit
from the adoption of such shared community infrastructure
to handle the mundane and time-consuming details of data interchange.
The \texttt{tskit} library (Section~\ref{sec-efficiency})
is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard may ease software implementation burden
on researchers, freeing their time to address the many fascinating open
questions and challenges that exist.

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
Peter Ralph and Andrew Vaughn
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
